The combination of reporting bias and underpowered study designs have substantially exaggerated the motor learning benefits of self-controlled practice and enhanced expectancies: A meta-analysis

### 2022-08-28 docs/preprint/preprint.tex
update: added fig1 and description of data sets

@@

In addition to providing an overall effect estimate, the RoBMA-PSMA model can also be used to calculate Bayes factors (BF) quantifying the strength of support for the presence (or absence) of an intervention effect, heterogeneity, or reporting bias. To illustrate, consider the question of whether there is or is not an intervention effect. Since we used a neutral prior that considered each possibility equally likely, we can take the ratio of the posterior probabilities of model ensembles that included an effect to those that did not. We always report the BFs so they can be interpreted as how many times more likely the data were assuming the best supported hypothesis compared to the other hypothesis. For example, a BF01 = 2.0 for the absence of an intervention effect suggests that the models assuming all variation is random fit the data twice as well as models predicting an effect is present\footnote{BF10 is the inverse of BF01, so while BF01 indicates evidence in support of the null hypothesis, BF10 indicates evidence in support of the alternative hypothesis.}. Bayes factors should be interpreted as a continuous measure of the relative support for one hypothesis versus another, not as the probability a hypothesis is true.

@@

Statistical analyses were conducted using R (Version 4.1.2; R Core Team, 2021) and the R-packages \emph{geomtextpath} (Version 0.1.0; Cameron \& van den Brand, 2022), \emph{gt} (Version 0.6.0; Iannone et al., 2022), \emph{invgamma} (Version 1.1; Kahle \& Stamey, 2017), \emph{metafor} (Version 3.4.0; Viechtbauer, 2010), \emph{papaja} (Version 0.1.0.9999; Aust \& Barth, 2020), \emph{patchwork} (Version 1.1.0.9000; Pedersen, 2022), \emph{plotly} (Version 4.10.0; Sievert, 2020), \emph{pwr} (Version 1.3.0; Champely, 2020), \emph{renv} (Version 0.15.5; Ushey, 2022), \emph{robma} (Bartoš \& Maier, 2020), \emph{tidyverse} (Version 1.3.1; Wickham et al., 2019), \emph{tinylabels} (Version 0.2.3; Barth, 2022), \emph{truncdist} (Version 1.0.2; Novomestky \& Nadarajah, 2016), and \emph{zcurve} (Version 2.1.2; Bartoš \& Schimmack, 2020) were used in this project.

@@

The model-averaged posterior distribution of the average effect from the RoBMA-PSMA model is displayed in Figure \ref{fig:fig1}B. The results suggest moderate evidence against the presence of an effect, BF01 = 3.16, very weak evidence against the presence of heterogeneity, BF01 = 1.7, and overwhelming evidence for the presence of reporting bias, BF10 = 18,399. The overall model ensemble estimated the effect of self-controlled practice as \emph{d} = .034 (95\% credible interval {[}.0, .248{]}). Heterogeneity was estimated as \(\tau\) = .05 (95\% credible interval {[}.0, .261{]}). A model fit with two influential cases included found overwhelming evidence for the presence of heterogeneity, BF10 = 1,924,516 and estimated \(\tau\) = .559 (95\% credible interval {[}.36, .78{]}). There were no other meaningful differences between models.

@@

The model-averaged posterior distribution of the average effect from the RoBMA-PSMA model is displayed in Figure \ref{fig:fig1}C. The results revealed weak evidence for the presence of an effect BF10 = 1.9, weak evidence for the presence of reporting bias, BF10 = 2.3, and very strong evidence for the presence of heterogeneity, BF10 = 47.6. The model ensemble estimated an average effect of \emph{d} = .26 (95\% credible interval {[}-.07, .63{]}). Heterogeneity was estimated as \(\tau\) = .35 (95\% credible interval {[}.07, .54{]}). A model fit with two influential cases included found weak evidence for the absence of an effect BF01 = 2.5, strong evidence for the presence of reporting bias, BF10 = 21, and overwhelming evidence for the presence of heterogeneity, BF10 = 45,300. The model ensemble estimated an average effect of \emph{d} = .00 (95\% credible interval {[}-.62, .59{]}), \(\tau\) = .49 (95\% credible interval {[}.32, .68{]}).

The RoBMA-PSMA models with and without influential cases differed primarily with respect to the fit of the PEESE models. When two large effect sizes with large standard errors were included in the analysis, the PEESE model provided a very strong fit to the data (BF10 = 39.9) when assuming no true effect but the presence of heterogeneity. When the two influential cases were excluded, the best fitting model was the PET under the same assumptions (BF10 = 9.4).

@@

\caption{\normalfont \small \textbf{Results from the robust Bayesian meta-analysis with publication selection model-averaging (RoBMA-PSMA) method.} \textbf{(A)} Prior spike and distribution (purple) with 50\% of prior probability density concentrated on the null hypothesis and 50\% reflecting plausible true effects with \emph{M} = 0 and \emph{SD} = 1. (The x-axis is truncated; the prior did include a small probability of effects larger than 1 and smaller than -1.) The model-averaged posterior distribution of the mean effect (Mu) of \textbf{(B)} self-controlled practice on motor learning (blue). The increased height of the spike at mu = 0 reflects increased belief in the null hypothesis. The remaining distribution reflects updated belief in the size and direction of a possible true effect. The mean estimate (\emph{M} = .034) for the posterior distribution is represented by the dashed vertical line. The model-averaged posterior distribution of the mean effect of \textbf{(C)} enhanced expectancies on motor learning (red). The decreased height of the spike at mu = 0 reflects decreased belief in the null hypothesis. The remaining distribution reflected updated belief in the size and direction of a possible true effect. The mean estimate (\emph{M} = .26) for the posterior distribution is represented by the dashed vertical line. The model-averaged posterior distribution generated from an analysis of \textbf{(D)} simulated data (green). In the simulation, 49 studies were sampled from a population with a true effect of mu = .54 and no reporting bias. The mean estimate (\emph{M} = .504) for the posterior is represented by the vertical dashed line. Outliers were not included in the self-controlled (\emph{n} = 2) and enhanced expectancies (\emph{n} = 2) models.}\label{fig:fig1}
\end{figure}

@@

The results of the \emph{z}-curve analysis can be seen in Figure \ref{fig:fig2}B. The analysis estimated the average statistical power (expected discovery rate) of all experiments examining the effect of self-controlled practice was 6\% (95\% confidence intervals {[}5\%, 12\%{]}). Since the upper bound of the expected discovery rate does not overlap with the observed discovery rate of 48\% (95\% confidence interval {[}35\%, 62\%{]}), there is evidence of significant reporting bias. The estimated conditional power of the statistically significant results (expected replication rate) was 11\% (95\% confidence interval {[}3\%, 30\%{]}). Including influential cases did not markedly change the results.

@@

The results of the \emph{z}-curve analysis of enhanced expectancy and self-controlled practice meta-data combined can be seen in Figure \ref{fig:fig2}D. The expected discovery rate of all studies conducted on the motivational factors in OPTIMAL theory is 6\% (95\% confidence interval {[}5\%, 13\%{]}). The observed discovery rate was 46\% (95\% confidence interval {[}37\%, 56\%{]}). The lower bound of the observed discovery rate does not overlap with the upper bound of the expected discovery rate, providing evidence of significant reporting bias in this literature. The expected replication rate of the statistically significant results was 21\% (95\% confidence interval {[}4\%, 39\%{]}). Including influential cases did not meaningfully change the results.

@@

@@

\caption{\normalfont \small \textbf{Results of the \emph{z}-curve analyses.} Distribution of \emph{z}-values for \textbf{(A)} our simulation with 47\% results and no reporting bias (green), \textbf{(B, left)} self-controlled practice (blue), \textbf{(C, left)} enhanced expectancies (red), and \textbf{(D, left)} motivational factors combined (purple). Values in the \emph{z}-score distributions for each analysis that are to the right of the significance line (\emph{z} = 1.96; solid, black) are statistically significant with a two-tailed \(\alpha\) of .05. Bootstrapped 95\% confidence distributions for the expected discovery rate (EDR; dark) and expected replication rate (ERR; light) for \textbf{(B, right)} self-controlled practice (blue), \textbf{(C, right)} enhanced expectancies (red), and \textbf{(D, right)} motivational factors combined (purple). The expected discovery rate is the estimated average power of all studies that have been conducted. The expected replication rate is the estimated power of all studies that observed a statistically significant result. The analysis estimated the average statistical power to be 6\%, 8\%, and 6\% for self-controlled practice, enhanced expectancies, and motivational factors combined, respectively. The estimated conditional power of the statistically significant results was 11\%, 33\%, and 21\% for self-controlled practice, enhanced expectancies, and motivational factors combined, respectively. Note that a minimum power of 80\% (dashed line, black) is often recommended. Outliers were not included in the self-controlled (\emph{n} = 2), enhanced expectancies (\emph{n} = 2), and motivational factors combined (\emph{n} = 4) analyses.}\label{fig:fig2}
\end{figure}

@@

Our objective in the current study was to conduct a holistic assessment of the motivation pillar of OPTIMAL theory. Specifically, we used RoBMA-PSMA---a state-of-the-art bias correction technique---to evaluate the effect of self-controlled practice and enhanced expectancies on motor learning (see Figure \ref{fig:fig1}). Additionally, we used \emph{z}-curve analyses to estimate the statistical power of the enhanced expectancies studies alone and then combined with the self-controlled practice studies to assess the motivation pillar of OPTIMAL theory (see Figure \ref{fig:fig2}). Using RoBMA-PSMA to model the effect of self-controlled practice on motor learning, we found moderate evidence against the presence of an effect (BF01 = 3.16). The model ensemble estimated a small average effect of \emph{d} = .034 with a 95\% credible interval that reached zero {[}.0, .248{]}. We also found overwhelming evidence for the presence of reporting bias (BF10 = 18,399). When modeling the effect of enhanced expectancies on motor learning with RoBMA-PSMA, we found weak evidence for the presence of an effect (BF10 = 1.9). The model ensemble estimated a small average effect of \emph{d} = .26 with a 95\% credible interval that included zero {[}-.07, .63{]}. Weak evidence for the presence of reporting bias (BF10 = 2.3) was also found. There was very strong support for the presence of heterogeneity in the enhanced expectancies literature (BF10 = 47.6).

@@

\noindent The first and last author would like to thank František Bartoš for drawing their attention to the \texttt{RoBMA} package during the review of the original self-controlled learning meta-analysis.

>>>>

\hypertarget{description-of-datasets}{%
\subsection{Description of datasets}\label{description-of-datasets}}

The original meta-analyses followed similar inclusion criteria and data extraction policies (see Figure \ref{fig:fig1}). All data were extracted by two independent researchers with conflicts settled by including a third researcher. The literature search ended in August of 2019 in the self-controlled practice meta-analysis, and 10 months later in June 2020 in the enhanced expectancies meta-analysis.

\caption{\normalfont \small \textbf{Inclusion criteria, dependent variable selection, studies screended and outcomes included in each meta-analysis.}}\label{fig:fig1}
\end{figure}

@@

In addition to providing an overall effect estimate, the RoBMA-PSMA model can also be used to calculate Bayes factors (BF) quantifying the strength of support for the presence (or absence) of an intervention effect, heterogeneity, or reporting bias. To illustrate, consider the question of whether there is or is not an intervention effect. Since we used a neutral prior that considered each possibility equally likely, we can take the ratio of the posterior probabilities of model ensembles that included an effect to those that did not. We always report the BFs so they can be interpreted as how many times more likely the data were assuming the best supported hypothesis compared to the other hypothesis. For example, a BF01 = 2.0 for the absence of an intervention effect suggests that the models assuming all variation is random fit the data twice as well as models predicting an effect is present\footnote{BF\textsubscript{10} is the inverse of BF\textsubscript{01}, so while BF\textsubscript{01} indicates evidence in support of the null hypothesis, BF\textsubscript{10} indicates evidence in support of the alternative hypothesis.}. Bayes factors should be interpreted as a continuous measure of the relative support for one hypothesis versus another, not as the probability a hypothesis is true.

@@

Statistical analyses were conducted using R (Version 4.2.1; R Core Team, 2021) and the R-packages \emph{geomtextpath} (Version 0.1.0; Cameron \& van den Brand, 2022), \emph{gt} (Version 0.6.0; Iannone et al., 2022), \emph{invgamma} (Version 1.1; Kahle \& Stamey, 2017), \emph{metafor} (Version 3.4.0; Viechtbauer, 2010), \emph{papaja} (Version 0.1.0.9999; Aust \& Barth, 2020), \emph{patchwork} (Version 1.1.0.9000; Pedersen, 2022), \emph{plotly} (Version 4.10.0; Sievert, 2020), \emph{pwr} (Version 1.3.0; Champely, 2020), \emph{renv} (Version 0.15.5; Ushey, 2022), \emph{robma} (Bartoš \& Maier, 2020), \emph{tidyverse} (Version 1.3.1; Wickham et al., 2019), \emph{tinylabels} (Version 0.2.3; Barth, 2022), \emph{truncdist} (Version 1.0.2; Novomestky \& Nadarajah, 2016), and \emph{zcurve} (Version 2.1.2; Bartoš \& Schimmack, 2020) were used in this project.

@@

The model-averaged posterior distribution of the average effect from the RoBMA-PSMA model is displayed in Figure \ref{fig:fig1}B. The results suggest moderate evidence against the presence of an effect, BF\textsubscript{01} = 3.16, very weak evidence against the presence of heterogeneity, BF\textsubscript{rf} = 1.7, and overwhelming evidence for the presence of reporting bias, BF\textsubscript{pb} = 18,399. The overall model ensemble estimated the effect of self-controlled practice as \emph{d} = .034 (95\% credible interval {[}.0, .248{]}). Heterogeneity was estimated as \(\tau\) = .05 (95\% credible interval {[}.0, .261{]}). A model fit with two influential cases included found overwhelming evidence for the presence of heterogeneity, BF\textsubscript{rf} = 1,924,516 and estimated \(\tau\) = .559 (95\% credible interval {[}.36, .78{]}). There were no other meaningful differences between models.

@@

The model-averaged posterior distribution of the average effect from the RoBMA-PSMA model is displayed in Figure \ref{fig:fig1}C. The results revealed weak evidence for the presence of an effect BF\textsubscript{10} = 1.9, weak evidence for the presence of reporting bias, BF\textsubscript{pb} = 2.3, and very strong evidence for the presence of heterogeneity, BF\textsubscript{rf} = 47.6. The model ensemble estimated an average effect of \emph{d} = .26 (95\% credible interval {[}-.07, .63{]}). Heterogeneity was estimated as \(\tau\) = .35 (95\% credible interval {[}.07, .54{]}). A model fit with two influential cases included found weak evidence for the absence of an effect BF\textsubscript{01} = 2.5, strong evidence for the presence of reporting bias, BF\textsubscript{pb} = 21, and overwhelming evidence for the presence of heterogeneity, BF\textsubscript{rf} = 45,300. The model ensemble estimated an average effect of \emph{d} = .00 (95\% credible interval {[}-.62, .59{]}), \(\tau\) = .49 (95\% credible interval {[}.32, .68{]}).

The RoBMA-PSMA models with and without influential cases differed primarily with respect to the fit of the PEESE models. When two large effect sizes with large standard errors were included in the analysis, the PEESE model provided a very strong fit to the data (BF\textsubscript{10} = 39.9) when assuming no true effect but the presence of heterogeneity. When the two influential cases were excluded, the best fitting model was the PET under the same assumptions (BF\textsubscript{10} = 9.4).

@@

\caption{\normalfont \small \textbf{Results from the robust Bayesian meta-analysis with publication selection model-averaging (RoBMA-PSMA) method.} \textbf{(A)} Prior spike and distribution (purple) with 50\% of prior probability density concentrated on the null hypothesis and 50\% reflecting plausible true effects with \emph{M} = 0 and \emph{SD} = 1. (The x-axis is truncated; the prior did include a small probability of effects larger than 1 and smaller than -1.) The model-averaged posterior distribution of the mean effect (Mu) of \textbf{(B)} self-controlled practice on motor learning (blue). The increased height of the spike at mu = 0 reflects increased belief in the null hypothesis. The remaining distribution reflects updated belief in the size and direction of a possible true effect. The mean estimate (\emph{M} = .034) for the posterior distribution is represented by the dashed vertical line. The model-averaged posterior distribution of the mean effect of \textbf{(C)} enhanced expectancies on motor learning (red). The decreased height of the spike at mu = 0 reflects decreased belief in the null hypothesis. The remaining distribution reflected updated belief in the size and direction of a possible true effect. The mean estimate (\emph{M} = .26) for the posterior distribution is represented by the dashed vertical line. The model-averaged posterior distribution generated from an analysis of \textbf{(D)} simulated data (green). In the simulation, 49 studies were sampled from a population with a true effect of mu = .54 and no reporting bias. The mean estimate (\emph{M} = .504) for the posterior is represented by the vertical dashed line. Outliers were not included in the self-controlled (\emph{n} = 2) and enhanced expectancies (\emph{n} = 2) models.}\label{fig:fig2}
\end{figure}

@@

The results of the \emph{z}-curve analysis can be seen in Figure \ref{fig:fig3}B. The analysis estimated the average statistical power (expected discovery rate) of all experiments examining the effect of self-controlled practice was 6\% (95\% confidence intervals {[}5\%, 12\%{]}). Since the upper bound of the expected discovery rate does not overlap with the observed discovery rate of 48\% (95\% confidence interval {[}35\%, 62\%{]}), there is evidence of significant reporting bias. The estimated conditional power of the statistically significant results (expected replication rate) was 11\% (95\% confidence interval {[}3\%, 30\%{]}). Including influential cases did not markedly change the results.

@@

The results of the \emph{z}-curve analysis of enhanced expectancy and self-controlled practice meta-data combined can be seen in Figure \ref{fig:fig3}D. The expected discovery rate of all studies conducted on the motivational factors in OPTIMAL theory is 6\% (95\% confidence interval {[}5\%, 13\%{]}). The observed discovery rate was 46\% (95\% confidence interval {[}37\%, 56\%{]}). The lower bound of the observed discovery rate does not overlap with the upper bound of the expected discovery rate, providing evidence of significant reporting bias in this literature. The expected replication rate of the statistically significant results was 21\% (95\% confidence interval {[}4\%, 39\%{]}). Including influential cases did not meaningfully change the results.

@@

{\centering \includegraphics{../../figs/fig3}

@@

\caption{\normalfont \small \textbf{Results of the \emph{z}-curve analyses.} Distribution of \emph{z}-values for \textbf{(A)} our simulation with 47\% results and no reporting bias (green), \textbf{(B, left)} self-controlled practice (blue), \textbf{(C, left)} enhanced expectancies (red), and \textbf{(D, left)} motivational factors combined (purple). Values in the \emph{z}-score distributions for each analysis that are to the right of the significance line (\emph{z} = 1.96; solid, black) are statistically significant with a two-tailed \(\alpha\) of .05. Bootstrapped confidence distributions for the expected discovery rate (EDR; dark) and expected replication rate (ERR; light) for \textbf{(B, right)} self-controlled practice (blue), \textbf{(C, right)} enhanced expectancies (red), and \textbf{(D, right)} motivational factors combined (purple). Note, reported CIs include 5 extra points (EDR) and 3 extra points (ERR) added to the quantiles of the bootstrapped distributions in the right panel, consistent with Bartoš et al. (2022). The expected discovery rate is the estimated average power of all studies that have been conducted . The expected replication rate is the estimated power of all studies that observed a statistically significant result. The analysis estimated the average statistical power to be 6\%, 8\%, and 6\% for self-controlled practice, enhanced expectancies, and motivational factors combined, respectively. The estimated conditional power of the statistically significant results was 11\%, 33\%, and 21\% for self-controlled practice, enhanced expectancies, and motivational factors combined, respectively. Note that a minimum power of 80\% (dashed line, black) is often recommended. Outliers were not included in the self-controlled (\emph{n} = 2), enhanced expectancies (\emph{n} = 2), and motivational factors combined (\emph{n} = 4) analyses.}\label{fig:fig3}
\end{figure}

@@

Our objective in the current study was to conduct a holistic assessment of the motivation pillar of OPTIMAL theory. Specifically, we used RoBMA-PSMA---a state-of-the-art bias correction technique---to evaluate the effect of self-controlled practice and enhanced expectancies on motor learning (see Figure \ref{fig:fig1}). Additionally, we used \emph{z}-curve analyses to estimate the statistical power of the enhanced expectancies studies alone and then combined with the self-controlled practice studies to assess the motivation pillar of OPTIMAL theory (see Figure \ref{fig:fig2}). Using RoBMA-PSMA to model the effect of self-controlled practice on motor learning, we found moderate evidence against the presence of an effect (BF\textsubscript{01} = 3.16). The model ensemble estimated a small average effect of \emph{d} = .034 with a 95\% credible interval that reached zero {[}.0, .248{]}. We also found overwhelming evidence for the presence of reporting bias (BF\textsubscript{pb} = 18,399). When modeling the effect of enhanced expectancies on motor learning with RoBMA-PSMA, we found weak evidence for the presence of an effect (BF\textsubscript{10} = 1.9). The model ensemble estimated a small average effect of \emph{d} = .26 with a 95\% credible interval that included zero {[}-.07, .63{]}. Weak evidence for the presence of reporting bias (BF\textsubscript{pb} = 2.3) was also found. There was very strong support for the presence of heterogeneity in the enhanced expectancies literature (BF\textsubscript{rf} = 47.6).

@@

\noindent The first and last author would like to thank František Bartoš for drawing their attention to the \texttt{RoBMA} package during the review of the original self-controlled learning meta-analysis and for providing helpful feedback on an earlier draft of this manuscript.
